---
title: "R : Sample Size and Power Calculations for Tests Comparing Means in Clinical Research"
author: "John Pauline Pineda"
date: "May 7, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    highlight: tango
    css: doc.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=15, fig.height=10)
```

# **1. Table of Contents**
|
| This document presents a non-exhaustive list of sample size and power calculations for clinical research mean comparison tests using various helpful packages in <mark style="background-color: #CCECFF">**R**</mark>. 
|
| Mean comparison tests applied during clinical research refer to trials evaluated in terms of mean responses of some primary study endpoints. The objectives of the intended clinical trials usually include the evaluation of intervention effects, demonstration of therapeutic equivalence or non-inferiority and establishment of superiority. The equations applied in this study (mostly contained in the <mark style="background-color: #CCECFF">**base**</mark> package) attempt to provide calculations for the sample size, critical value and power based on the study objectives and hypotheses of interest. 
|
##  1.1 One-Sample Design

###  1.1.1 Test for Equality

```{r section_1.1.1, warning=FALSE, message=FALSE}

```

###  1.1.2 Test for Non-Inferiority

```{r section_1.1.2, warning=FALSE, message=FALSE}

```

###  1.1.3 Test for Superiority

```{r section_1.1.3, warning=FALSE, message=FALSE}

```

###  1.1.4 Test for Equivalence

```{r section_1.1.4, warning=FALSE, message=FALSE}

```

##  1.2 Two-Sample Parallel Design

###  1.2.1 Test for Equality

```{r section_1.2.1, warning=FALSE, message=FALSE}

```

###  1.2.2 Test for Non-Inferiority

```{r section_1.2.2, warning=FALSE, message=FALSE}

```

###  1.2.3 Test for Superiority

```{r section_1.2.3, warning=FALSE, message=FALSE}

```

###  1.2.4 Test for Equivalence

```{r section_1.2.4, warning=FALSE, message=FALSE}

```

##  1.3 Two-Sample Crossover Design

###  1.3.1 Test for Equality

```{r section_1.3.1, warning=FALSE, message=FALSE}

```

###  1.3.2 Test for Non-Inferiority

```{r section_1.3.2, warning=FALSE, message=FALSE}

```

###  1.3.3 Test for Superiority

```{r section_1.3.3, warning=FALSE, message=FALSE}

```

###  1.3.4 Test for Equivalence

```{r section_1.3.4, warning=FALSE, message=FALSE}

```

##  1.4 Multiple-Sample One-Way ANOVA Design

###  1.4.1 Pairwise Comparison

```{r section_1.4.1, warning=FALSE, message=FALSE}

```

###  1.4.2 Simultaneous Comparison

```{r section_1.4.2, warning=FALSE, message=FALSE}

```

##  1.5 Multiple-Sample Williams Design

###  1.5.1 Test for Equality

```{r section_1.5.1, warning=FALSE, message=FALSE}

```

###  1.5.2 Test for Non-Inferiority

```{r section_1.5.2, warning=FALSE, message=FALSE}

```

###  1.5.3 Test for Superiority

```{r section_1.5.3, warning=FALSE, message=FALSE}

```

###  1.5.4 Test for Equivalence

```{r section_1.5.4, warning=FALSE, message=FALSE}

```

# **2. References**
|
| **[Book]** [Sample Size Calculations in Clinical Research](https://www.taylorfrancis.com/books/mono/10.1201/9781315183084/sample-size-calculations-clinical-research-shein-chung-chow-jun-shao-hansheng-wang-yuliya-lokhnygina) by Shein-Chung Chow, Jun Shao, Hansheng Wang and Yuliya Lokhnygina
| **[Book]** [Clinical Trial Data Analysis Using R and SAS](https://www.taylorfrancis.com/books/mono/10.1201/9781315155104/clinical-trial-data-analysis-using-sas-ding-geng-din-chen-karl-peace-pinggao-zhang) by Ding-Geng Chen, Karl Peace and Pinggao Zhang
| **[Book]** [Design and Analysis of Experiments with R](https://www.taylorfrancis.com/books/mono/10.1201/b17883/design-analysis-experiments-john-lawson) by John Lawson
| **[Book]** [Design and Analysis of Experiments](https://link.springer.com/book/10.1007/978-3-319-52250-0) by Angela Dean , Daniel Voss and Danel Draguljić
| **[Book]** [Design and Analysis of Experiments](https://bcs.wiley.com/he-bcs/Books?action=index&bcsId=10790&itemId=1119320933) by Douglas Montgomery
| **[Article]** [Power and Sample Size](https://powerandsamplesize.com/) by HyLown Consulting Team
| **[Article]** [An Introduction to Different Types of Study Design](https://s4be.cochrane.org/blog/2021/04/06/an-introduction-to-different-types-of-study-design/) by Hadi Abbas
| **[Article]** [Case-Control and Cohort Studies: A Brief Overview](https://s4be.cochrane.org/blog/2017/12/06/case-control-and-cohort-studies-overview/) by Saul Crandon
| **[Article]** [Cohort Studies: Prospective and Retrospective Designs](https://s4be.cochrane.org/blog/2019/03/06/cohort-studies-prospective-retrospective-designs/) by Izabel de Oliveira
| **[Article]** [Prevalence vs. Incidence: What is the Difference](https://s4be.cochrane.org/blog/2020/11/06/prevalence-vs-incidence-what-is-the-difference/) by Georgina Ford
| **[Article]** [Randomized Clinical Trial (RCT): Simple Definition, Phases, and Types](https://www.statisticshowto.com/experimental-design/randomized-clinical-trial-rcts/) by Statistics-How-To Team
|
|
|
|