---
title: "R : Sample Size and Power Calculations for Tests Comparing Means in Clinical Research"
author: "John Pauline Pineda"
date: "May 7, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    highlight: tango
    css: doc.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=15, fig.height=10)
```

# **1. Table of Contents**

##  1.1 One-Sample Design

###  1.1.1 Test for Equality

```{r section_1.1.1, warning=FALSE, message=FALSE}

```

###  1.1.2 Test for Non-Inferiority

```{r section_1.1.2, warning=FALSE, message=FALSE}

```

###  1.1.3 Test for Superiority

```{r section_1.1.3, warning=FALSE, message=FALSE}

```

###  1.1.4 Test for Equivalence

```{r section_1.1.4, warning=FALSE, message=FALSE}

```

##  1.2 Two-Sample Parallel Design

###  1.2.1 Test for Equality

```{r section_1.2.1, warning=FALSE, message=FALSE}

```

###  1.2.2 Test for Non-Inferiority

```{r section_1.2.2, warning=FALSE, message=FALSE}

```

###  1.2.3 Test for Superiority

```{r section_1.2.3, warning=FALSE, message=FALSE}

```

###  1.2.4 Test for Equivalence

```{r section_1.2.4, warning=FALSE, message=FALSE}

```

##  1.3 Two-Sample Crossover Design

###  1.3.1 Test for Equality

```{r section_1.3.1, warning=FALSE, message=FALSE}

```

###  1.3.2 Test for Non-Inferiority

```{r section_1.3.2, warning=FALSE, message=FALSE}

```

###  1.3.3 Test for Superiority

```{r section_1.3.3, warning=FALSE, message=FALSE}

```

###  1.3.4 Test for Equivalence

```{r section_1.3.4, warning=FALSE, message=FALSE}

```

##  1.4 Multiple-Sample One-Way ANOVA Design

###  1.4.1 Pairwise Comparison

```{r section_1.4.1, warning=FALSE, message=FALSE}

```

###  1.4.2 Simultaneous Comparison

```{r section_1.4.2, warning=FALSE, message=FALSE}

```

##  1.5 Multiple-Sample Williams Design

###  1.5.1 Test for Equality

```{r section_1.5.1, warning=FALSE, message=FALSE}

```

###  1.5.2 Test for Non-Inferiority

```{r section_1.5.2, warning=FALSE, message=FALSE}

```

###  1.5.3 Test for Superiority

```{r section_1.5.3, warning=FALSE, message=FALSE}

```

###  1.5.4 Test for Equivalence

```{r section_1.5.4, warning=FALSE, message=FALSE}

```

# **2. References**
|
| **[Book]** [Sample Size Calculations in Clinical Research](https://www.taylorfrancis.com/books/mono/10.1201/9781315183084/sample-size-calculations-clinical-research-shein-chung-chow-jun-shao-hansheng-wang-yuliya-lokhnygina) by Shein-Chung Chow, Jun Shao, Hansheng Wang and Yuliya Lokhnygina
| **[Book]** [Clinical Trial Data Analysis Using R and SAS](https://www.taylorfrancis.com/books/mono/10.1201/9781315155104/clinical-trial-data-analysis-using-sas-ding-geng-din-chen-karl-peace-pinggao-zhang) by Ding-Geng Chen, Karl Peace and Pinggao Zhang
| **[Book]** [Design and Analysis of Experiments with R](https://www.taylorfrancis.com/books/mono/10.1201/b17883/design-analysis-experiments-john-lawson) by John Lawson
| **[Book]** [Design and Analysis of Experiments](https://link.springer.com/book/10.1007/978-3-319-52250-0) by Angela Dean , Daniel Voss and Danel Draguljić
| **[Book]** [Design and Analysis of Experiments](https://bcs.wiley.com/he-bcs/Books?action=index&bcsId=10790&itemId=1119320933) by Douglas Montgomery
|
|
|
|